---
title: "Impactors of Heart Disease"
author: "Marcela, Xinyi, Pia"
date: "4/26/2022"
output: 
  html_document:
    df_print: paged
    code_download: true
    theme: cerulean
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,warning = FALSE,message=FALSE)
```

```{r}
library(scales)
library(tidyverse)
library(gt)
library(lubridate)     
library(patchwork)
library(ggthemes)      
library(janitor)       
library(fmsb)
library(RColorBrewer) 
library(ggthemes)    
library(shiny)        
library(knitr)
library(rsconnect)
library(ggtext)
theme_set(theme_minimal()) 
```

```{r,echo=FALSE}
heart<-read_csv("heart_2020_cleaned.csv")
```

# INTRODUCTION & BACKGROUND

Heart disease is one of the leading causes of death for most Americans, regardless of race (CDC). Around 47% of Americans have at least 1 out of 3 major risk factors for heart disease, which are: high cholesterol, high blood pressure, and smoking. Obesity (High BMI), diabetes, physical activity levels, and alcohol consumption are also considered as key indicators. Identifying the factors that have the most influence on heart disease prevalence is important in order to act on them earlier. By looking at data from CDC’s Behavioral Risk Factor Surveillance System (BRFSS), we aim to detect patterns that could potentially help predict a patient’s predisposition to heart disease. 

Our group chose to focus our research on those main factors: BMI, Race, Sex, Age, and some lifestyle choices such as smoking, drinking, and sleeptime. 


## DATA COLLECTION

As mentioned, the dataset originates from the CDC and the Behavioral Risk Factor Surveillance System (BRFSS). The annual survey, which is conducted through phone calls, collects information on the health of residents in the USA.It was established in 1984 with 15 states and now gathers data from all 50 states as well as the District of Columbia and three U.S. territories. It is one of the largest consistently performed health survey in the world, with more than 400,000 adult participants per year. The data we are using in this project comes from the 2020 survey of 400,000 adults in the country, which is the most recent publicly shared version form the CDC. 


## ANALYSIS

***Demographic Break-down of Sample***

The following three tables provide a quick look at the demographic breakdown of the sample from our dataset. The tables include the number and percentage of survey participants according to their Race, Sex, and Age.

```{r,echo=FALSE}
heart1 <- heart %>% 
  count(Sex) %>% 
  mutate(Percentage = (n/sum(n))) %>% 
  rename(Count = n)

heart_gt <- heart1 %>% 
  gt(rowname_col = "row",
    groupname_col = "group") %>% 
  tab_options(column_labels.background.color = "lightblue") %>% 
   tab_header(title = "Sex of Sample from Dataset",
    subtitle = md("")
  ) %>% 
  
  fmt_integer(columns = Count) %>% 
  fmt_percent(columns = Percentage)

heart_gt

```

\  


```{r,echo=FALSE}
race <- heart %>% 
  count(Race) %>% 
  mutate(Percentage = (n/sum(n))) %>% 
  rename(Count = n)

race_gt <- race %>% 
  gt(rowname_col = "row",
    groupname_col = "group") %>% 
  tab_options(column_labels.background.color = "lightpink") %>% 
   tab_header(
    title = "Race of Sample from Dataset",
    subtitle = md("")
  ) %>% 
  fmt_integer(columns = Count) %>% 
  fmt_percent(columns = Percentage)

race_gt
```
\  

```{r, echo=FALSE}
age <- heart %>% 
  count(AgeCategory) %>% 
  mutate(Percentage = (n/sum(n))) %>% 
  rename(Count = n,
         Ages = AgeCategory) 

age_gt <- age %>% 
  gt(rowname_col = "row",
    groupname_col = "group") %>% 
  tab_options(column_labels.background.color = "lightgreen") %>% 
   tab_header(
    title = "Age Ranges of Sample from Dataset",
    subtitle = md("")
  ) %>% 
  fmt_integer(columns = Count) %>% 
  fmt_percent(columns = Percentage)

age_gt
```
\  



***Heart Disease Prevalence According to Demographics***

**I. Race, Age Group, Sex, and the Likelihood of Heart Disease**
\  

```{r,echo=FALSE}
gt1 <- heart %>%
  group_by(HeartDisease) %>%
  count(Sex) %>%
  mutate(Percentage = 100*(n/sum(n))) %>%
  ggplot(aes(y = Sex,
             x = n,
             fill = HeartDisease)) +
  geom_col() +
  scale_x_continuous(labels = label_number(suffix = "K", scale = .001)) +
  scale_fill_manual(values = c("#008B8B","#7FFFD4"))+
  labs(title = "Heart Disease Prevalence by Sex") +
  theme(
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank(),
        plot.title = element_text(hjust = 0.5,face="bold",size=10),
        plot.caption = element_text(size = 8.5),
        legend.position = "bottom")

gt2 <- heart %>%
  mutate(Race= str_replace(Race, "American Indian/Alaskan Native", "American Indian or
                               Alaskan Native")) %>%
  group_by(Race) %>%
  count(HeartDisease) %>%
  ggplot(aes(x= n, y = fct_reorder(Race, n), fill = HeartDisease)) +
  geom_col() +
  scale_x_continuous(labels = label_number(suffix = "K", scale = .001)) +
  labs(title = "Heart Disease Prevalence by Race") +
  scale_fill_manual(values = c("darkslateblue","deepskyblue3"))+
  theme(axis.text.y.left = element_text(size = 7),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank(),
        plot.title = element_text(hjust = 0.5,face="bold",size=10),
        plot.caption = element_text(size = 8),
        legend.position = "bottom")

gt3 <- heart %>%
  mutate(Race= str_replace(Race, "American Indian/Alaskan Native", "American Indian or
                               Alaskan Native")) %>%
  group_by(AgeCategory) %>%
  count(HeartDisease) %>%
  ggplot(aes(y = AgeCategory,
             x = n,
             fill = HeartDisease), position = "stack") +
  geom_col() +
  scale_x_continuous(labels = label_number(suffix = "K", scale = .001)) +
  labs(title = "Heart Disease Prevalence by Age Group") +
  scale_fill_manual(values = c("#8A2BE2","#FFB6C1"))+
  theme(axis.text.y.left = element_text(size = 7),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank(),
        plot.title = element_text(hjust = 0.5,face="bold",size=10),
        plot.caption = element_text(size = 8),
        legend.position = "bottom", 
        )

gt3 + (gt2/gt1) 
```



Heart disease is primarily found within age groups over 50. 20% of individuals older than 50 reported heart disease, accounting for one-fifth of the population. Meanwhile, less than 5% of survey participants under the age of 50 report heart disease. In each age group over 50 but under 75, around 10% to 16% erport heart disease. 
\  
Given that most of the sample is white, the overall rate of heart disease prevalence for white identifying folks seems to be overwhelmingly high. This means that these findings cannot be generalized to the represent larger populations outside of the survey. 
\  
Males appear to have higher rates of heart disease, with about 11% reporting it, than females do at about 6.5%.




**II. BMI and Heart Disease**

BMI stands for Body Mass Index. BMIs are values taken from the mass and height of an individual. One is able to calculate their BMI by taking body mass (kg) and dividing it by the square of body height (m).


Below are the different categories of BMIs:


  Underweight = <18.5
  
  Normal wright = 18.5-24.9
  
  Overweight = 25-29.9
  
  Obesity Class I = 30-34.9
  
  Obesity Class 2 = 35-35.9
  
  Extreme Obesity = >40
  
  
  \  
  
```{r,echo=FALSE}
bmi_groups <- heart %>% 
  mutate(bmi_group=case_when(BMI<18.5~"Underweight",
                         BMI>=18.5 & BMI<=24.9~"Normal Weight",
                         BMI>=25 & BMI<=29.9~"Overweight",
                         BMI>=30 & BMI<=34.9~"Obesity Class I",
                         BMI>=35 & BMI<=35.9~"Obesity Class II",
                         BMI>=36~"Extreme Obesity"))
heart %>% 
  filter(BMI<50) %>% 
  ggplot(aes(x=HeartDisease,y=BMI))+
  geom_boxplot(fill = "cornflowerblue", 
               alpha = .7)+
  labs(title="BMI Average of Individuals With and Without Heart Disease")+
  theme(plot.title = element_text(hjust=0.5,face="bold"))
```
\  

The graph above shows that subjects with heart disease have slightly higher BMI than subjects without heart disease. This indicates that higher BMIs mean an increased possibility of experiencing heart disease. 


\  




```{r,echo=FALSE}
bmi_groups %>% 
  filter(is.na(bmi_group)==FALSE) %>% 
  ggplot(aes(x=HeartDisease,fill=bmi_group))+
  geom_bar(stat="count",width=0.5,position = "dodge")+
  geom_text(stat='count',aes(label=scales::percent(..count../sum(..count..),accuracy=0.1)), color="black", size=3.5, vjust=-0.5)+
  facet_wrap(vars(bmi_group))+
  labs(title="BMI groups with and without Heart Disease",
       y="")+
  scale_fill_manual(values = c("darkslateblue","deepskyblue3","#8B4513","#DB7093","#8B008B","#4682B4"))+
  scale_y_continuous(expand = expansion(mult = c(0,0.1)))+
  theme(legend.position="none",
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank(),
        panel.grid.major.x = element_blank(),
        plot.title = element_text(hjust=0.5,face="bold"))
```
\  

In this dataset, there is a disproportionate amount of survey participants that fall under the BMI category of overweight. It may therefore be a stretch to conclude that overweight subjects, as well as those classified under normal and obese, are less likely to have heart disease.


\  



```{r,echo=FALSE}
bmi_groups %>% 
  group_by(Race,HeartDisease,Sex) %>% 
  summarise(mean_bmi=mean(BMI)) %>% 
  ggplot(aes(x=HeartDisease,y=mean_bmi,fill=Sex))+
  geom_bar(stat = "identity",position="dodge")+
  geom_text(aes(label = round(mean_bmi,2)),position = position_dodge(.9), vjust = 1.5, colour = "white", size = 2)+
  scale_y_continuous(labels=scales::comma)+
  facet_wrap(vars(Race))+
  scale_fill_manual(values = c("darkslateblue","deepskyblue3"))+
  labs(title="Heart Disease According to the BMI Average Among Different Races and Sexes",
       y="Average BMI") +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        plot.title = element_text(face = "bold",
                                  size = 10))
```

\  

There is no significant distinction between males and females in regards to reporting heart disease. This implies that sex is not an key indicator of Heart Disease. However, males generally have relatively higher BMI averages than female do. 
\  
Across all race and sex groups, survey participants that reported heart disease generally have higher BMI averages, further strengthening the hypothesis that higher BMI increases the likelihood of experiencing heart disease.
\  


**IV. Heart Disease and Lifestyle Choices**

A. Sleep time

```{r,echo=FALSE}
bmi_groups %>% 
  ggplot(aes(x=SleepTime,fill=HeartDisease))+
  geom_histogram()+ 
  facet_wrap(vars(HeartDisease),scales="free_y",ncol=1)+
  scale_fill_manual(values=c('#EEE8AA','#B8860B'))+
  labs(title="Heart Disease Correlated to Average Sleep Time",
       x="Average Sleep Time(hour)")+
  theme(plot.title = element_text(face="bold",color="black",hjust=0.5),
        legend.position = "none",
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank())

```




The average sleeptime of participants with and without heart disease don't vary by much, with most of the sample reporting sleeptime at around 6-8 hours a day. However, we can observe a notable amount of individuals sleeping over 10 hours and reporting heart disease, which may connect to the scientific theory that oversleeping is correlated to Cardiovascular disease.

\  


B. Heart Disease when Drinking Alcohol and Smoking
```{r,echo=FALSE}
drinking<-bmi_groups %>% 
  ggplot(aes(x=AlcoholDrinking,fill=HeartDisease))+
  geom_bar(stat="count",width=0.5,position='fill')+
  scale_fill_manual(values=c('#999999','#E69F00'))+
  geom_text(stat='count',aes(label=scales::percent(..count../sum(..count..))), color="white", size=3.2,position=position_fill(0.5))+
  labs(y="",x="",
       title = "Heart Disease Prevalence\nBased on Drinking Habits")+
  theme(axis.text.y = element_blank(),
        plot.title = element_text(face = "bold",
                                  size = 12,
                                  hjust = 0.5),
        legend.position = "bottom", 
        panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank())
  

smoking<-bmi_groups %>% 
  ggplot(aes(x=Smoking,fill=HeartDisease))+
  theme(axis.text.y = element_blank())+
   geom_bar(stat="count",width=0.5,position='fill')+
  geom_text(stat='count',aes(label=scales::percent(..count../sum(..count..))), color="white", size=3.2,position=position_fill(0.5))+
  labs(y="",x="",
title = "Heart Disease Prevalence\nBased on Smoking Habits")+
theme(axis.text.y = element_blank(),
        plot.title = element_text(face = "bold",
                                  size = 12,
                                  hjust = 0.5),
      legend.position = "bottom",
      panel.grid.major.x = element_blank(),
      panel.grid.major.y = element_blank(),
      panel.grid.minor.x = element_blank(),
      panel.grid.minor.y = element_blank()
        )

drinking+smoking
```



Approximately 92% of survey participants who consume alcohol do not report heart disease, suggesting that alcohol consumption is not a deciding factor. Further supporting this theory, the percentage of people without heart disease in the non-drinking group is at 91.4%.
\  
In the non-smoking group, 93%  reported no heart disease. Meanwhile, there is a lower percentage of participants in the smoking group (87%) that did not report heart disease. This could suggest thaht smoking habits have a bigger impact on heart disease than drinking.
\  



**V. Heart Disease Calculator**

Using this calculator tool, individuals may indicate characteristics and lifestyle choices to see the proportion of participants from the survey with similar backgrounds who have and do not have heart disease. Through comparisons, they may be able to have some insight on the likelihood of getting heart disease. 


```{r, echo=FALSE, show_col_types = FALSE}
 #Reading the dataset 

health <- read_csv("heart_2020_cleaned.csv")

#Data
bmi_groups <- health %>% 
  clean_names() %>% 
  mutate(bmi_group=case_when(bmi<18.5~"Underweight",
                             bmi>=18.5 & bmi<=24.9~"Normal Weight",
                             bmi>=25 & bmi<=29.9~"Overweight",
                             bmi>=30 & bmi<=34.9~"Obesity Class I",
                             bmi>=35 & bmi<=35.9~"Obesity Class II",
                             bmi>=36~"Extreme Obesity"))
```



```{r, echo=FALSE, show_col_types = FALSE}
shinyApp(

ui = fluidPage(
  sidebarLayout(
  sidebarPanel(
    numericInput(inputId = "weight",
               label = "Weight (Kg)",
               value = "65"),
  numericInput(inputId = "height",
               label = "Height (Cm)",
               value = "170"),
  selectInput(inputId = "sex",
               label = "Sex",
               choices = list(Female = "Female", Male = "Male")),
  selectInput(inputId = "age_cat",
               label = "Age Category",
               choices = bmi_groups %>% 
                 arrange(age_category) %>% 
                 distinct(age_category) %>% 
                 pull(age_category),
               multiple=FALSE),
  selectInput(inputId = "phys_act",
               label = "Physical Activity",
               choices = list(Yes = "Yes", No = "No")),
  selectInput(inputId = "smoking",
              label = "Smoke",
              choices = list(Yes = "Yes", No = "No")),
  submitButton(text="Ready!")),
  mainPanel(
    plotOutput(outputId = "Percentage_plot")
  ))),


server=function(input,output){
 
  output$Percentage_plot <- renderPlot({
   
    
    weight <- input$weight
    height <- input$height
    
    dataset <- tibble(weight,
                      height) %>%
      mutate(BMI_val = weight/((height/100)^2),
             bmi_cat = case_when(BMI_val<18.5~"Underweight",
                                 BMI_val>=18.5 & BMI_val<25~"Normal Weight",
                                 BMI_val>=25 & BMI_val<30~"Overweight",
                                 BMI_val>=30 & BMI_val<35~"Obesity Class I",
                                 BMI_val>=35 & BMI_val<36~"Obesity Class II",
                                 BMI_val>=36~"Extreme Obesity")) 
    user_bmi <- dataset$bmi_cat
    
    app_data <- bmi_groups %>%
      filter(bmi_group == user_bmi,
             sex == input$sex,
             age_category == input$age_cat,
             physical_activity == input$phys_act,
             smoking == input$smoking) %>% 
      group_by(heart_disease) %>% 
      mutate(count_per_HDPrevalence= n()) %>% 
      ungroup() %>% 
      mutate(total= n(),
             prop= (count_per_HDPrevalence/total))
  
    
    bar_distribution <- app_data%>%
      ggplot()+
      geom_bar(aes(x=1, fill = heart_disease), position = "fill",show.legend = FALSE)+
      theme_clean()
    
    pie <- bar_distribution +
      coord_polar("y", start= 0)+
      scale_fill_brewer(palette="Pastel1")+
      theme_void()+
      geom_text(aes(x= 1, y= prop/1.5, label= scales::percent(prop, accuracy = 0.01)))+
      labs(x= NULL,
           y= NULL,
           title= "Percentage of Population of Similar Characteristics <span style = 'color: lightblue2;'>with Heart Disease</span>",
           caption = "User manual:\n 
           Please input your weight in kilograms and height in centimeters.\n Select your age category and whether you exercise and/or smoke.\n
           Click on the button once you are ready to find out the percentage of surveyed\n people of similar characteristics to those indicated\n by you who suffer from heart disease.")+
      theme(plot.title = element_markdown())

    pie
  })
}
)
```

\  

## Conclusion

In this project, we hope to shift individuals’ priorities towards prevention rather than medical intervention.By visualizing data from the CDC and Behavioral Risk Factor Surveillance System (BRFSS), readers will have the ability to better understand information that may otherwise be difficult to comprehend or access — information that can also be critical to the decision-making processes that people undergo on a daily basis. By seeing what characteristics and activities make one more vulnerable to experiencing heart disease, individuals can hopefully make the necessary adjustments to their lifestyles if necessary. At the end of the day, the people have the power to choose what they want to do with the information that is given to them; however, having sufficient and accurate sources are necessary for people to make well-informed choices. 

